hbm_paper_results

### 2024-01-04 paper_results.Rmd
assumption testing for w/in feat sex-diffs in error t testing

#also load raw data
ukb_cn_df <- fread("/Users/megardn/Desktop/BGD_Repos/combat_biovar/data/ukb_to_model/ukb_CN_data_simsites.csv")
```

@@

```{r}
#define function
pheno_diff.list <- paste0("abs.diff_", pheno_list)

centile.t.tests <- function(df, feature_list=pheno_diff.list, comp_multiplier=1){
  #initialize empty dfs to store outputs
  t.df <- data.frame("pheno" = character(),
                     "group1" = character(),
                     "group2" = character(),
                     "p.value" = double())
  #run tests
  attach(df)
  for (pheno in feature_list) {
    df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=df[[pheno]], g=df[["dataset"]], paired = TRUE, p.adj = "none"))
    df.pairwise.t$pheno <- pheno
    t.df <- rbind(t.df, df.pairwise.t)
  }
  n_comp <- nrow(t.df)

@@

@@

Within each phenotype, iterate paired t test of ranks between each combo of combat dfs (nonparametric per discussion with Andrew). (see Zimmerman & Zumbo, 1993). Using custom functions from `ranked_welch_tests.R`.

```{r}
#initalize empty dfs to store outputs
sum.df <- data.frame(pheno = character(),
                     term = character(),
                     df = double(),
                     sumsq = double(),
                     meansq = double(),
                     statistic = double(),
                     p.value = double())
t.df <- data.frame("pheno" = character(),
                   "group1" = character(),
                   "group2" = character(),
                   "p.value" = double())

#run tests
attach(diffs.df)
for (pheno in pheno_diff.list) {
  # df.aov <- tidy(kruskal.test(diffs.df[[pheno]] ~ diffs.df[["dataset"]]))
  # df.aov$pheno <- pheno
  # sum.df <- rbind(sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=diffs.df[[pheno]], g=diffs.df[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()

#apply fdr correction
# sum.df <- sum.df %>%
#   dplyr::filter(term != "Residuals") %>%
#   mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(pheno_list)))

t.df <- t.df %>%
  mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(pheno)),
         sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                             p.val_fdr >= 0.05 ~ FALSE),
         pheno = sub("abs.diff_", "", pheno)) %>%
  unite(comp, c("group1", "group2")) %>%
  mutate(comp = as.factor(comp))

t.df.sum <- t.df %>%
  group_by(comp) %>%
  summarize(n_sig_regions = sum(sig_fdr),
            n_compare = n(),
            min_pval = min(p.val_fdr),
            max_pval = max(p.val_fdr))
t.df.sum
```

@@

```{r}
subj_sum.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, fill=sex)) +
  labs(y= "Magnitude Centile Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat configurations")

subj_sum.df %>%
  mutate(mean_cent_diff = mean_cent_diff*100) %>% #for visuals
  ggplot(aes(x=dataset, y=mean_cent_diff, fill=sex)) +
  geom_boxplot(trim=FALSE, scale="width") +
  labs(y="Centile Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean error in centile scores across ComBat configurations")

subj_sum.df %>%
  dplyr::filter(dataset == "cf.gamlss" | dataset == "cf.gam" | dataset == "cf.lm") %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot(aes(x=sim.site, y=mean_cent_abs.diff, fill=sex)) +
  geom_violin(trim=FALSE) +
  facet_wrap(~dataset) +
  labs(y="Magnitude Centile Error", x = "Simulated Site") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude error in centile scores across ComBat configurations")

@@

@@

#add more info from parcellations
dk.parc <- dk$data %>%
  as.data.frame() %>%
  na.omit() %>%
  dplyr::select(c(hemi, region, label)) %>%
  distinct()

@@

```{r}
#within each phenotype, iterate paired t test between each combo of combat dfs 
pheno_diff.list.z <- paste0("abs.diff_", pheno_list, ".z")

z.t.tests <- function(df){
  #initialize empty dfs to store outputs
  t.df <- data.frame("pheno" = character(),
                     "group1" = character(),
                     "group2" = character(),
                     "p.value" = double())
  #run tests
  attach(df)
  for (pheno in pheno_diff.list.z) {
    tryCatch(df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=df[[pheno]], g=df[["dataset"]], paired = TRUE, p.adj = "none")), error=function(e) { traceback() }) #
    df.pairwise.t$pheno <- pheno
    t.df <- rbind(t.df, df.pairwise.t)
  }
  n_comp <- nrow(t.df)

>>>>

#also load raw data
ukb_cn_df <- fread("/Users/megardn/Desktop/BGD_Repos/combat_biovar/data/ukb_to_model/ukb_CN_data_simsites.csv")

#plotting info from dk atlas
dk.parc <- dk$data %>%
  as.data.frame() %>%
  na.omit() %>%
  dplyr::select(c(hemi, region, label)) %>%
  distinct()
```

@@

```{r}
#define function
pheno_abs.diff.list <- paste0("abs.diff_", pheno_list)

centile.t.tests <- function(df, feature_list=pheno_abs.diff.list, comp_multiplier=1){
  #initialize empty dfs to store outputs
  t.df <- data.frame("pheno" = character(),
                     "group1" = character(),
                     "group2" = character(),
                     "p.value" = double())
  #run tests
  attach(df)
  for (pheno in feature_list) {
    df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=df[[pheno]], g=df[["dataset"]], paired = TRUE, p.adj = "none"))
    df.pairwise.t$pheno <- pheno
    t.df <- rbind(t.df, df.pairwise.t)
  }
  n_comp <- nrow(t.df)

@@

modifying to test w/in feature rather than mean centiles

```{r}
sex.bias.feat.t.tests <- function(df, feature_list=pheno_list, comp_multiplier=1){

  #initialize empty df to store outputs
  t.df <- data.frame("dataset" = character(),
                     "pheno" = character(),
                     "p.value" = double(),
                     "sex_diff" = double(),
                     "t.stat" = double(),
                     "df" = double())
  attach(df)
  for (pheno in feature_list) {
  #conduct t test
  df.sex.t <- df %>%
    group_by(dataset) %>%
    summarise(p.value = tidy(rank.welch.t.test.formula(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$p.value,
              sex_diff = tidy(rank.welch.t.test.formula(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$estimate,
              t.stat = tidy(rank.welch.t.test.formula(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$statistic,
              df = tidy(rank.welch.t.test.formula(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$parameter) %>%
    ungroup()
  df.sex.t$pheno <- pheno
  t.df <- rbind(t.df, df.sex.t)
  }

  #apply fdr correction
  n_comp=length(unique(df$dataset))*length(feature_list)
  df.sex.t <- t.df %>%
    dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = (n_comp*comp_multiplier)), #use comp_multiplier if using fun with lapply
           sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                               p.val_fdr >= 0.05 ~ FALSE),
           dataset=as.factor(dataset))

  #summarize
  df.sex.t.sum <- df.sex.t %>%
    group_by(dataset) %>%
    dplyr::summarize(n_sig_regions = sum(sig_fdr),
              min_pval = min(p.val_fdr),
              max_pval = max(p.val_fdr)) %>%
    ungroup() %>%
    as.data.frame()

  return(df.sex.t.sum)
  #detach()
}

trying to get interpretable effect sizes w t.test (rather than ranked t test)

```{r}
sex.bias.feat.t.tests_2 <- function(df, feature_list=pheno_list, comp_multiplier=1){

  #initialize empty df to store outputs
  t.df <- data.frame("dataset" = character(),
                     "pheno" = character(),
                     "p.value" = double(),
                     "sex_diff" = double(),
                     "t.stat" = double(),
                     "df" = double())
  attach(df)
  for (pheno in feature_list) {
  #conduct t test
  df.sex.t <- df %>%
    group_by(dataset) %>%
    summarise(p.value = tidy(t.test(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$p.value,
              sex_diff = tidy(t.test(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$estimate,
              t.stat = tidy(t.test(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$statistic,
              df = tidy(t.test(formula=as.formula(paste0("diff_",pheno, " ~ sex")), paired = FALSE, p.adj = "none"))$parameter) %>%
    ungroup()
  df.sex.t$pheno <- pheno
  t.df <- rbind(t.df, df.sex.t)
  }

  #apply fdr correction
  n_comp=length(unique(df$dataset))*length(feature_list)
  df.sex.t <- t.df %>%
    dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = (n_comp*comp_multiplier)), #use comp_multiplier if using fun with lapply
           sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                               p.val_fdr >= 0.05 ~ FALSE),
           dataset=as.factor(dataset))

  return(df.sex.t)
  #detach()
}

@@

Do centile scores change based on combat version?
Foregoing anova based on emails with andrew.
Within each phenotype, iterate paired t test of ranks between each combo of combat dfs (nonparametric per discussion with Andrew). (see Zimmerman & Zumbo, 1993). Using custom functions from `ranked_welch_tests.R`.

```{r}
centile.t.tests(diffs.df)
```

@@

```{r}
subj_sum.df %>%
  dplyr::filter(dataset == "cf.gamlss" | dataset == "cf.gam" | dataset == "cf.lm") %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_boxplot(aes(x=sex, y=mean_cent_abs.diff, fill=sex)) +
  labs(y= "Magnitude Centile Error", x = "ComBat Configurations") +
  facet_wrap(~dataset) +
  theme(legend.position="none") +
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat configurations")

subj_sum.df %>%
  mutate(mean_cent_diff = mean_cent_diff*100) %>% #for visuals
  ggplot(aes(x=dataset, y=mean_cent_diff, fill=sex)) +
  geom_violin(trim=FALSE, scale="width") +
  labs(y="Centile Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean error in centile scores across ComBat configurations")

subj_sum.df %>%
  dplyr::filter(dataset == "cf.gamlss" | dataset == "cf.gam" | dataset == "cf.lm") %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot(aes(x=sim.site, y=mean_cent_abs.diff, fill=sex)) +
  geom_violin(trim=FALSE) +
  facet_wrap(~dataset) +
  labs(y="Magnitude Centile Error", x = "Simulated Site") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude error in centile scores across \nComBat configurations")

@@

Sig test w/in feature diffs

check for skew (welch t.test vs ranked welch t test)
```{r}
for (pheno in pheno_list){
  val <- paste0("diff_", pheno)
  p <- ggplot(diffs.df, aes(x = diffs.df[[val]], fill= sex, color=sex)) +
      geom_histogram(alpha=0.5, position="identity") +
    ggtitle(val)
print(p)
}
```

```{r}
sex.bias.feat.t.tests(diffs.df)
```

Sig test w/in feature diffs
```{r, warning=FALSE}
sex_t_tests_in_feat.df <- sex.bias.feat.t.tests_2(diffs.df)
```

```{r}
sex_t_tests_in_feat.df.sum <- sex_t_tests_in_feat.df %>%
    group_by(dataset) %>%
    dplyr::summarize(n_sig_regions = sum(sig_fdr),
              min_pval = min(p.val_fdr),
              max_pval = max(p.val_fdr)) %>%
    ungroup() %>%
    as.data.frame()
```

```{r}
sex_t_tests_in_feat.df2 <- sex_t_tests_in_feat.df %>%
  mutate(pheno_cat = as.factor(case_when(
    pheno %in% vol_list_global ~ "Global Volume",
    pheno %in% vol_list_regions ~ "Regional Volume",
    pheno %in% sa_list ~ "Regional SA",
    pheno %in% ct_list ~ "Regional CT",
    TRUE ~ NA_character_)),
    label = sub("_[^_]*_", "_", pheno)) #for plotting
sex_t_tests_to_plot <- left_join(sex_t_tests_in_feat.df2, dk.parc, by="label")
```

plot w/in feature t tests

```{r}
#var plot
sex.var.plot.df <- sex_t_tests_to_plot %>%
  dplyr::filter(pheno_cat != "Global Volume") %>%
  dplyr::filter(!grepl('_refA', dataset)) %>% #drop refA levels
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE),
         sex_diff = case_when(sig_fdr == FALSE ~ NA,
                              TRUE ~ sex_diff)) %>%
  group_by(dataset, pheno_cat)

tp <- unique(sex.var.plot.df[,c('dataset','pheno_cat')]) %>%
  dplyr::filter(dataset=="ukb_CN")
tp$sex_diff <- 1

sex.var.plot <- ggplot(sex.var.plot.df) +
  geom_rect(data = tp, fill="gray", xmin = -Inf,xmax = Inf,
            ymin = -Inf, ymax = Inf, alpha = 0.4) +
  ggseg::geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = sex_diff)) +
  #scale_fill_gradient2() +
  facet_grid(dataset ~ pheno_cat, labeller = labeller(dataset = as_labeller(dataset_names))) + 
  theme_linedraw(base_size=12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA), aspect.ratio=2/3, plot.margin = margin(1,1,1,1, 'pt')) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork", name = "Sex Effect", limits = c(-0.003, 0.003))
sex.var.plot

```{r}
#plot global
sex.var.plot.global.df <- sex_t_tests_to_plot %>%
  dplyr::filter(pheno_cat == "Global Volume") %>%
  dplyr::filter(!grepl('_refA', dataset)) %>% #drop refA levels
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  dplyr::select(!c(label, region, hemi)) %>% #drop cols corresponding to dk regions
  mutate(tissue_class = as.factor(case_when(pheno == "GMV" ~ "cGM",
                                            pheno == "Ventricles" ~ "CSF",
                                            pheno == "WMV" ~ "WM",
                                            pheno == "sGMV" ~ "sGM",
                                            TRUE ~ NA)),
        sex_diff = case_when(sig_fdr == FALSE ~ NA,
                              TRUE ~ sex_diff)) %>% #drop NS values
  group_by(dataset)

tp <- unique(sex.var.plot.global.df[,c('dataset','pheno_cat')]) %>%
  dplyr::filter(dataset=="ukb_CN")
tp$sex_standard <- 1

sex.var.plot.global <- ggplot(sex.var.plot.global.df) +
  geom_rect(data = tp, fill="gray", xmin = -Inf,xmax = Inf,
            ymin = -Inf, ymax = Inf, alpha = 0.4) +
  ggseg::geom_brain(atlas = aseg_cerebrum, 
             aes(fill = sex_diff),
             side = c("coronal")) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_linedraw(base_size=12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA), aspect.ratio=, plot.margin = margin(1,1,1,1, 'pt'), strip.text.y = element_blank()) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork", name = "Sex Effect", limits = c(-0.003, 0.003))
sex.var.plot.global
```

@@

@@

```{r}
#within each phenotype, iterate paired t test between each combo of combat dfs 
pheno_abs.diff.list.z <- paste0("abs.diff_", pheno_list, ".z")

z.t.tests <- function(df){
  #initialize empty dfs to store outputs
  t.df <- data.frame("pheno" = character(),
                     "group1" = character(),
                     "group2" = character(),
                     "p.value" = double())
  #run tests
  attach(df)
  for (pheno in pheno_abs.diff.list.z) {
    tryCatch(df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=df[[pheno]], g=df[["dataset"]], paired = TRUE, p.adj = "none")), error=function(e) { traceback() }) #
    df.pairwise.t$pheno <- pheno
    t.df <- rbind(t.df, df.pairwise.t)
  }
  n_comp <- nrow(t.df)
